---
title: "R Notebook"
output: html_notebook
---


```{r libraries, include=FALSE}
library(tidyverse)
library(rredlist)
```

```{r read & organize data}
amphibians <- read.csv("data/redlist_Amphibians_IDName.csv") %>% 
  distinct(BINOMIAL, .keep_all = TRUE) %>% 
  mutate(taxon = "amph") %>% 
  select(origId = ID_NO, sciName = BINOMIAL, taxon)
birds <- read.csv("data/redlist_Birds_IDName.csv") %>% 
  distinct(SCINAME, .keep_all = TRUE) %>% 
  mutate(taxon = "bird") %>% 
  select(origId = SPCRECID, sciName = SCINAME, taxon)
fish <- read.csv("data/redlist_Fish_IDName.csv") %>% 
  distinct(BINOMIAL, .keep_all = TRUE) %>% 
  mutate(taxon = "fish", .keep_all = TRUE) %>% 
  select(origId = HShedID, sciName = BINOMIAL, taxon)
reptiles <- read.csv("data/redlist_Mammals_IDName.csv") %>% 
  distinct(binomial, .keep_all = TRUE) %>% 
  mutate(taxon = "rptl") %>% 
  select(origId = id_no, sciName = binomial, taxon)
mammals<- read.csv("data/redlist_Reptiles_IDName.csv") %>% 
  distinct(binomial, .keep_all = TRUE) %>% 
  mutate(taxon = "maml") %>% 
  select(origId = id_no, sciName = binomial, taxon)

allSpecies <- amphibians %>% bind_rows(birds) %>% bind_rows(fish) %>%
  bind_rows(reptiles) %>% bind_rows(mammals) # %>% 
  # head(10)
```

Prepare the CSV file to write the results into. If it doesn't exist, create it.

Since the web API loop takes a couple of hours to complets, it is susceptible to random internet issues that might cause it to abort. So had to figure out a way to run for all species in incremental steps. So, if the results file exists, we make sure the web API loop runs only for species it doesn't already have.

```{r}
resultFile <- "results/redlistRead.csv"
fileColumns <- "origId, sciName, taxon, year, category, code"

if (!file.exists(resultFile)) {
  file.create(resultFile)
  write_lines(fileColumns, resultFile)
  speciesCompleted <- 0
} else {
  completed <- read.csv(resultFile) %>% 
    distinct(sciName, .keep_all = TRUE)
  speciesCompleted <- nrow(completed)
}

if (speciesCompleted == 0) {
  remainingSpecies <- allSpecies
} else {
  remainingSpecies <- anti_join(allSpecies, completed, by = "sciName")
}

print(paste(nrow(allSpecies), "species in total."), quote = FALSE)
print(paste(nrow(remainingSpecies), "species to go."), quote = FALSE)
if (nrow(remainingSpecies) > 0) {
  pb = txtProgressBar(min = 0, max = nrow(remainingSpecies), initial = 0, style = 3) 
  for (i in 1:nrow(remainingSpecies)) {
    speciesEntry <- remainingSpecies[i,]
    assessmentHistory <- rl_history(speciesEntry$sciName)
    Sys.sleep(3)
    if (is_empty(assessmentHistory$result)) {
      assessmentHistory$result <- data.frame(
        year = NA, code = NA, category = NA)
    }
    appended <- assessmentHistory$result %>% 
      mutate(
        origId = first(speciesEntry$origId),
        taxon = first(speciesEntry$taxon),
        sciName = first(speciesEntry$sciName)) %>% 
      select(origId, sciName, taxon, year, category, code)
    write_csv(appended, resultFile, append = TRUE)
    setTxtProgressBar(pb,i)
  }
}

print("All species completed.", quote = FALSE)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
