---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Changes in IUCN Red List status from 2010 to present

Use the status history data for each species to tell whether its status has improved or worsened.

## Load libraries

```{r load libraries}
library(tidyverse)
```

# Load and reorganize status history data 

```{r load data}
assessmentHistories <- read.csv("results/redlistRead_complete.csv") %>% 
  drop_na(year) %>% 
  mutate(code = factor(code, ordered = TRUE),
         category = factor(category, ordered = TRUE)) %>% 
  mutate(
    cat_reassigned = fct_collapse(
      category,
      criticallyEndangered = "Critically Endangered",
      endangered = "Endangered",
      threatened = "Threatened",
      nearThreatened = "Near Threatened",
      vulnerable = "Vulnerable",
      leastConcern = "Least Concern",
      other = c("Rare", "Not Recognized", "Lower Risk/near threatened", "Lower Risk/least concern",
                "Lower Risk/conservation dependent", "Insufficiently Known", "Indeterminate",
                "Data Deficient", "\"Very rare but believed to be stable or increasing\"",
                "\"Very rare and believed to be decreasing in numbers\"",
                "\"Less rare but believed to be threatened-requires watching\"")
    ) %>% fct_relevel("criticallyEndangered", "endangered", "threatened",
                      "nearThreatened", "vulnerable", "leastConcern", "other")
  )
```
## Find change in status from 2010 to present


### Status in 2010

```{r status 2010}
pre2010Status <- assessmentHistories %>% 
  head(500) %>%
  filter(year <= 2010) %>% 
  group_by(sciName) %>% 
  summarise(
    status2010 = .data$cat_reassigned[.data$year == max(year)],
    status2010AssessmentYear = max(year)
  )
```

```{r status present}
withPreAndPost2010Status <- left_join(pre2010Status, assessmentHistories, by = "sciName") %>% 
  filter(year > 2010) %>% 
  group_by(sciName) %>% 
  summarise(
    status2010 = status2010,
    status2010AssessmentYear = status2010AssessmentYear,
    statusNow = .data$cat_reassigned[.data$year == max(year)],
    statusNowAssessmentYear = max(year)
  )
```

```{r}
withPreAndPost2010Status <- withPreAndPost2010Status %>% 
  mutate(
    statusWorsened = statusNow < status2010,
    statusImproved = statusNow > status2010
  ) %>% 
  distinct(sciName, .keep_all = TRUE)
```

## Session info

```{r}
sessionInfo()
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
