---
title: "Western Ghats species: Changes in IUCN Red List status from 2010 to now"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document: default
---

Use the status history data for each species and calculate the percentage change in number of endangered species from 2010 to now.

This includes amphibian, reptile, fish, mammal and bird species list Srini shared with us.

## Load libraries

```{r load libraries, message=FALSE}
library(tidyverse)
library(rredlist)
```

# Load and reorganize status history data 

Before proceeding: 

* Species which don't appear in the Red List return `NA`, those are dropped 
    + If no synonyms exist in the database, such species have `synonymousSpeciesUsed = 'none available'`
    + If synonyms exist but none of them appear in the Red List either, then such species have `synonymousSpeciesUsed = 'all available'`; there are no such species in our list.
* We are only interested in changes between "endangered" and non-"endangered" status, but the IUCN Red List web API returns several more categories in status histories. Those categories are appropriately collapsed and ordered before proceeding.
* Threat level ordering is according to [IUCN's criteria](https://www.iucnredlist.org/about/faqs#What%20are%20the%20Red%20List%20Categories%20and%20Criteria). Here we consider threat level `"vulnerable"` or higher as "endangered", and we calculate the percentage change in number of those.

We first check for how many of the species are `'extinct'` and which one/s. Turns out it is one species, _Raorchestes travancoricus_, which was thought to be extinct but was later rediscovered.

```{r load data}
rawData <- read.csv("results/redlistRead_complete_synonymRuns.csv")
extinctSpecies <- rawData %>% drop_na(year) %>% 
  filter(str_detect(str_to_lower(category), 'extinct')) %>% 
  select(sciName)
rawData %>% drop_na(year) %>% 
  filter(sciName == extinctSpecies$sciName) %>% 
  select(sciName, taxon, year, category, synonymousSpeciesUsed)
```

Categories of threat status are collapsed to fewer levels and ordered as desired, according to [IUCN's criteria](https://www.iucnredlist.org/about/faqs#What%20are%20the%20Red%20List%20Categories%20and%20Criteria).

```{r reorganize threat categories}
assessmentHistories <- rawData %>% 
  drop_na(year) %>% 
  mutate(category = factor(category, ordered = TRUE)) %>% 
  mutate(
    cat_reassigned = fct_collapse(
      category,
      extinct = "Extinct",
      criticallyEndangered = "Critically Endangered",
      endangered = c("Endangered", "Threatened"),
      nearThreatened = "Near Threatened",
      vulnerable = "Vulnerable",
      leastConcern = "Least Concern",
      other = c("Rare", "Not Recognized", "Lower Risk/near threatened", "Lower Risk/least concern",
                "Lower Risk/conservation dependent", "Insufficiently Known", "Indeterminate",
                "Data Deficient", "\"Very rare but believed to be stable or increasing\"",
                "\"Very rare and believed to be decreasing in numbers\"",
                "\"Less rare but believed to be threatened-requires watching\"")
    ) %>% 
      fct_relevel("other", "leastConcern", "nearThreatened", "vulnerable",
                  "endangered", "criticallyEndangered", "extinct")
  )
```
# Find change in status from 2010 to now

Assessments happen for different species at different times, so there is no fixed schedule of Red List updation. Hence, for our purposes, we choose the year 2010 status as the one prior, and closest, in time to it. We choose the most recent status as the status now.

For species that were assessed before 2010 and not after, we count them to have had their status unchanged.

Some species (`r rawData %>% filter(is.na(year)) %>% nrow()` of them) that didn't have valid status histories (including after trying their synonyms) will be dropped from this change claculations.

```{r list species without valid status histories}
rawData %>% filter(is.na(year)) %>% select(-origId, -code)
```

## Filter to species assessed before 2010

Percentage change in number of endangered species from 2010 to now can be calculated over only those species that have been assessed before 2010. Hence, first we find species that have been assessed before 2010
.
```{r assessed before 2010}
assessedBefore2010 <- assessmentHistories %>% filter(year <= 2010) %>% 
  distinct(sciName)

withFullAssessmentHistory <- left_join(
  assessedBefore2010,
  assessmentHistories,
  by = "sciName"
)

print(
  paste(
    nrow(withFullAssessmentHistory %>% distinct(sciName)),
    ": Num. species assessed before 2010"
  ),
  quote = FALSE
)
print(
  paste(
    nrow(assessmentHistories %>% distinct(sciName)),
    ": Total number of valid species in our dataset"
  ),
  quote = FALSE
)
```


## Find status in 2010

It is the status of the species at the maximum assessment year before 2010 (i.e., most recent year before 2010).

```{r status 2010, message=FALSE}
pre2010Status <- withFullAssessmentHistory %>% 
  filter(year <= 2010) %>% 
  group_by(sciName) %>% 
  summarise(
    status2010 = .data$cat_reassigned[.data$year == max(year)],
    status2010AssessmentYear = max(year)
  )
```

## Find status now

It is the status of the species at the most recent assessment year. For species which haven't been assessed since 2010, their status now is taken to be their status at 2010. It follows that such species are counted as having their status unchanged from 2010 to now.

```{r status now, message=FALSE}
nowStatus <- withFullAssessmentHistory %>% 
  group_by(sciName) %>% 
  summarise(
    statusNow = .data$cat_reassigned[.data$year == max(year)],
    statusNowAssessmentYear = max(year)
  )

withPreAndPost2010Status <- left_join(pre2010Status, nowStatus, by = "sciName")
```

A tally of species that were assessed before 2010 but not since, which have been counted as having their status unchanged since 2010:

```{r status now is from before 2010}
print(
  paste(
    nrow(withPreAndPost2010Status %>% 
           filter(statusNowAssessmentYear <= 2010)),
    ": Number of species assessed before 2010 but not since"
  ),
  quote = FALSE
)
```

## Find percentage change in status (2010 - now)

For assessment in 2010 and now, label species as endangered if their assessed threat level is `"vulnerable"` or higher ([see IUCN reference](https://www.iucnredlist.org/about/faqs#What%20are%20the%20Red%20List%20Categories%20and%20Criteria)). Count them and calculate percentage change in total numbers as follows:

$$
\text{Percentage change} = 100 \times\frac{N_{endangered \ now}  - N_{endangered \ @ \ 2010} }{N_{endangered \ @ \ 2010}} 
$$ 

### Check status ordering
The order of status level factors, from low to high, is (following [IUCN's ordering](https://www.iucnredlist.org/about/faqs#What%20are%20the%20Red%20List%20Categories%20and%20Criteria)):

```{r display status order}
min(withPreAndPost2010Status$status2010)
min(withPreAndPost2010Status$statusNow)
```

### Calculate percentage change

```{r percent change calculation}
endangeredBefore2010 <- withPreAndPost2010Status %>% 
  mutate(isEndangered = status2010 >= levels(.data$status2010)[4]) %>% 
  filter(isEndangered == TRUE)
numEndangeredBefore2010 <- endangeredBefore2010 %>% 
  nrow()
endangeredNow <- withPreAndPost2010Status %>% 
  mutate(isEndangered = statusNow >= levels(.data$statusNow)[4]) %>% 
  filter(isEndangered == TRUE)
numEndangeredNow <- endangeredNow %>% 
  nrow()

print(
  paste(
    100 * (numEndangeredNow - numEndangeredBefore2010) / numEndangeredBefore2010,
    ": Percentage change in number of endangered species among birds, mammals, reptiles, fish, amphibians"
  ),
  quote = FALSE
)
```

# Database citation & details of system and libraries used
Citation of the IUCN dataset version, main packages used:
```{r}
rl_citation()
citation("rredlist")
citation("tidyverse")
```

System details and libraries used:
```{r package details}
sessionInfo()
```
